---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SUMnlmr

<!-- badges: start -->
<!-- badges: end -->

The goal of SUMnlmr is to allow investigations of potentially non-linear 
relationships between an exposure and an outcome via a Mendelian randomization
framework, without requiring full access to individual level genetic data.

It is based on the existing package for individual data by James Staley: nlmr 
(available from  https://github.com/jrs95/nlmr ). 

The core concept is to split the process into two distinct halves: one requiring 
individual level data, which is converted into a semi-summarized form 
(create_nlmr_summary) by dividing the population into strata based on the 
IV-free exposure. Associations with the exposure and the outcome are estimated 
in each stratum.
In the second half, this semi-summarized form can then be shared, without 
compromising patient privacy, and investigated separately using two IV methods:
a fractional polynomial method (frac_poly_summ_mr) and a piecewise linear method
(piecewise_summ_mr). Both methods calculate a localised causal effect (LACE). 
The piecewise method fits a continuous piecewise linear function to these 
estimates, while the fractional polynomial method fits the best 1 or 2 term 
fractional polynomial.


## Functions

- *create_nlmr_summary* — prepares individual level data into semi-summarised
  form, ready to fit nlmr models. Supports three stratification methods:
  - `"residual"` — the original IV-free residual approach (Staley & Burgess, 2017)
  - `"ranked"` — the doubly-ranked method (Tian et al., 2022)
  - `"interaction"` — an extension of the ranked method that corrects for
    GxE-induced bias by removing the interaction component from the exposure
    before stratification (Zhou et al., 2026)
- *frac_poly_summ_mr* — performs IV analysis using fractional polynomials.
- *piecewise_summ_mr* — performs IV analysis using a piecewise linear function.

## Installation

You can install the released version of SUMnlmr from 
[GitHub](https://github.com/) with:

```{r}
# install.packages("devtools")
devtools::install_github("amymariemason/SUMnlmr")
```

## Example 1: Summarizing data

This is a basic example which shows you how to create the semi-summarized data 
form. First we create some practice data:

```{r example}
#devtools::load_all()
library(SUMnlmr)
## create some data to practise on
test_data<-create_ind_data(N=10000, beta2=2, beta1=1)
# this creates quadratic.Y  = x + 2x^2 + errorY 
round(head(test_data), 2)
```

Then we use create_nlmr_summary to summarise it.

```{r example2}

## create the summarized form
## 
summ_data<-create_nlmr_summary(y = test_data$quadratic.Y,
                                x = test_data$X,
                                g = test_data$g,
                                covar = NULL,
                                family = "gaussian",
                                strata_method = "residual", 
                                controlsonly = FALSE,
                                q = 10)

round(head(summ_data$summary), 2)
```

Note 1: This example uses the residual method, which relies on parametric assumptions of linearity and homogeneity in the instrument-exposure model. We do not believe these conditions are met in practice and do not recommend applying this method. We include it here for completeness.


If we have covariates we want to adjust for in our analysis, we need to include
them at this stage. 

```{r example3}
set.seed(9941)
## create the summarized form
summ_covar<-create_nlmr_summary(y = test_data$quadratic.Y,
                                x = test_data$X,
                                g = test_data$g,
                                covar = matrix(data=c(test_data$linear.Y,                                     test_data$sqrt.Y),ncol=2),
                                family = "gaussian",
                                strata_method = "residual", 
                                q = 10)
  
round(head(summ_covar$summary), 2)
```



Note 2: Because the covariates are included as a matrix, lm cannot detect factor variables and create automatic dummy variables for them. The easiest way to include factor variables is to make these dummy variables by hand instead using the [model.matrix](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/model.matrix.html)
command. e.g.

```{r example3b}
set.seed(7778)
## create a factor
test_data$centre<- as.factor(rbinom(nrow(test_data),4, 0.5))

#turn factor into binary contrasts against first factor
dummies<- model.matrix(~centre,data=test_data)[,2:5]

summ_covar2<-create_nlmr_summary(y = test_data$quadratic.Y,
                                 x = test_data$X,
                                 g = test_data$g,
                                 covar = dummies,
                                 family = "gaussian",
                                 q = 10)

round(head(summ_covar2$summary), 2)
```

These have used a single genetic variant count, but the method works identically with a genetic score function for g instead. 

Logistic or Cox models in the G-Y relationship can be used by changing the family option - see details in the create_nlmr_summary function description.  

### Relaxing the parametric assumption: the `"ranked"` method

It is also possible to implement the doubly-ranked method described in [Tian et al., 2022](https://www.biorxiv.org/content/10.1101/2022.06.28.497930v1). This replaces the parametric assumptions with a rank-preserving assumption. This is currently the default method of the package, but there are concerns about this method's application, particularly in UK Biobank, with thanks to Hamilton et al., 2023 (<https://doi.org/10.1007/s10654-024-01113-9>) for their examples of this. In particular, this method performs poorly when there is a GxE interaction, as explained in Zhao et al., 2026 (<https://doi.org/10.64898/2026.01.22.26344640>).


To illustrate, we simulate a population with two unobserved subgroups: a
low-exposure group where the instrument is weak (`bx` ≈ 0.1) and a
high-exposure group where the instrument is strong (`bx` ≈ 0.8):

```{r example3c}

set.seed(5172)
N <- 10000

subgroup <- rbinom(N, 1, 0.5)   # unobserved; 50/50 split
rk_g     <- rbinom(N, 2, 0.3)
rk_u     <- rnorm(N)

rk_X <- ifelse(subgroup == 0,
  1 + 0.1 * rk_g + 0.3 * rk_u + abs(rnorm(N, 0, 0.5)),   # low-exposure group
  6 + 0.8 * rk_g + 0.3 * rk_u + abs(rnorm(N, 0, 0.5)))   # high-exposure group
rk_Y <- rk_X + 2 * rk_X^2 + 0.8 * rk_u + rnorm(N)

# Residual method — fits one linear g coefficient for the whole population
summ_rk_resid <- create_nlmr_summary(
  y = rk_Y, x = rk_X, g = rk_g,
  family = "gaussian", strata_method = "residual", q = 10
)

# Ranked method — makes no assumption about the shape of the g-x relationship
summ_rk_ranked <- create_nlmr_summary(
  y = rk_Y, x = rk_X, g = rk_g,
  family = "gaussian", strata_method = "ranked", q = 10
)
```

The ranked method correctly shows low `bx` in the low-exposure strata (where
the instrument is weak) and high `bx` in the high-exposure strata. The residual
method, forced to use a single average coefficient, produces distorted
intermediate values throughout:

```{r example3c_compare}

data.frame(
  stratum     = 1:10,
  bx_residual = round(summ_rk_resid$summary$bx, 2),
  bx_ranked   = round(summ_rk_ranked$summary$bx, 2)
)
```

### Correcting for GxE interactions: the `"interaction"` method

This is an extension of the ranked method developed specifically to mitigate GxE-induced bias in that method, as detailed in Zhou et al., 2026 (<https://doi.org/10.64898/2026.01.22.2634464>). Stratification methods can be biased when a covariate modifies the effect of the instrument on the exposure (a GxE interaction). The `"interaction"` method tries to address this by estimating and removing the interaction component from the exposure before stratification.

Two additional matrices are supplied alongside the usual `covar` argument:

- `gxe_interaction` (**H**): the effect modifier(s) that interact with `g`
- `gxe_covar` (**F**, optional): additional covariates for the interaction model
  that appear as main effects only

The stratification model fitted internally is:
**X = β₀ + β_g·g + β_F·F + β_H·H + β_{g×H}·(g×H)**

The corrected exposure **X − β_{g×H}·(g×H)** is then used to form strata via
the ranked method. Within-stratum genetic associations are estimated using
`covar` as usual.

To illustrate the difference, we simulate data where `modifier` strongly
amplifies the effect of `g` on `X` (GxE coefficient = 1.5, compared to a main
g effect of 0.25):

```{r example3d}

set.seed(8341)
N <- 10000

gxe_g        <- rbinom(N, 2, 0.3)
gxe_modifier <- rnorm(N)
gxe_u        <- runif(N, 0, 1)

# X has a strong GxE: the effect of g on X scales with modifier
gxe_X <- 2 + gxe_g * (0.25 + 1.5 * gxe_modifier) + gxe_u + rexp(N, 1)
gxe_Y <- gxe_X + 2 * gxe_X^2 + 0.8 * gxe_u + rnorm(N)

# Ranked method — no correction for the GxE
summ_gxe_ranked <- create_nlmr_summary(
  y = gxe_Y, x = gxe_X, g = gxe_g,
  family        = "gaussian",
  strata_method = "ranked",
  q = 10
)

# Interaction method — corrects for g x modifier before stratifying
summ_gxe_interaction <- create_nlmr_summary(
  y               = gxe_Y,
  x               = gxe_X,
  g               = gxe_g,
  gxe_interaction = matrix(gxe_modifier, ncol = 1),
  family          = "gaussian",
  strata_method   = "interaction",
  q = 10
)
```

In the ranked method, `bx` sweeps dramatically across strata (even turning negative in the lowest strata) because the strata are confounded with the modifier. The interaction method recovers approximately constant `bx` values close to the true main effect of 0.25:

```{r example3d_compare}

data.frame(
  stratum     = 1:10,
  bx_ranked      = round(summ_gxe_ranked$summary$bx, 2),
  bx_interaction = round(summ_gxe_interaction$summary$bx, 2)
)
```

Once your data is in this format, the output data frame is all you need to share to fit the fractional polynomial or piecewise linear models onto the data.

## Example 2: Fitting a fractional polynomial model

Your data needs to be in the semi-summarised form as shown above. We can then
fit a fractional polynomial model:

```{r example4}


model<- with(summ_data$summary, frac_poly_summ_mr(bx=bx,
                  by=by, 
                  bxse=bxse, 
                  byse=byse, 
                  xmean=xmean,
                  family="gaussian",
                  fig=TRUE)
)


summary(model)


```
This also produces a graph of the fit with 95% confidence intervals.
This is a ggplot object and can be adjusted with ggplot commands
```{r example5}
library(ggplot2)
f <- function(x) (x + 2*x^2 - mean(summ_data$summary$xmean) -
                    2*mean(summ_data$summary$xmean)^2 )

plot1 <- model$figure+ 
  stat_function(fun = f, colour = "green") +
  ggtitle("fractional polynomial fit from semi-summarized data")

plot1
```
There are also p-values provided in p_test and p_het. This is identical to the 
testing provided by the nlmr package. First, testing the heterogeneity of the LACE estimates:

- `fp_d1_d2`: test between the fractional polynomial degrees
- `fp`: fractional polynomial non-linearity test
- `quad`: quadratic test
- `Q`: Cochran Q test

And also testing the heterogeneity of the genetic-exposure associations:

- `Q`: Cochran Q heterogeneity test
- `trend`: trend test

```{r example6}
model$p_tests

model$p_heterogeneity
```

## Example 3: Piecewise linear model

We can instead fit a piecewise linear model to the same summarised data

```{r plexample}

model2 <-with(summ_data$summary, piecewise_summ_mr(by, bx, byse, bxse, xmean, xmin,xmax, 
                  ci="bootstrap_se",
                  nboot=1000, 
                  fig=TRUE,
                  family="gaussian",
                  ci_fig="ribbon")
)

summary(model2)


```
Again the figure is a ggplot object and can be adjusted similarly. 

```{r pl2}
plot2 <- model2$figure+ 
  stat_function(fun = f, colour = "green") +
  ggtitle("piecewise linear fit from semi-summarized data")

plot2
```

## Example 4: Binary outcome

The functions above can also fit binary outcome data, via a generalised linear 
model. 

```{r bin}

test_data$y.bin<-stats::rbinom(size=1, p=0.5, n=10000)

# create summ data
summ_bin<-create_nlmr_summary(y = test_data$y.bin,
                                x = test_data$X,
                                g = test_data$g,
                                covar = NULL,
                                family = "binomial",
                                q = 10)

# fit fractional poly model


model3<- with(summ_bin$summary,frac_poly_summ_mr(bx=bx,
                  by=by, 
                  bxse=bxse, 
                  byse=byse, 
                  xmean=xmean,
                  family="binomial",
                  fig=TRUE)
)

summary(model3)


```
Not unsurprisingly, we find no evidence of an effect, causal or otherwise, 
as the binary outcome was randomly distributed. 

Looking instead at the semi-summarised UK Biobank datasets on LDL-cholesterol 
and CAD — one with and one without covariates — we can see a potentially 
non-linear trend in the univariate data, which becomes a clear linear trend once
covariates are included.

```{r bi2}

# fit piecewise linear model
model4 <-with(LDL_CAD, piecewise_summ_mr(by, bx, byse, bxse, xmean, xmin,xmax, 
                  ci="bootstrap_se",
                  nboot=1000, 
                  fig=TRUE,
                  family="gaussian",
                  ci_fig="ribbon")
)


summary(model4)


# fit piecewise linear model
model5 <-with(LDL_CAD_covar,piecewise_summ_mr(by, bx, byse, bxse, xmean, xmin,xmax, 
                  ci="bootstrap_se",
                  nboot=1000, 
                  fig=TRUE,
                  family="gaussian",
                  ci_fig="ribbon")
)

summary(model5)

```
