---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SUMnlmr

<!-- badges: start -->
<!-- badges: end -->

The goal of SUMnlmr is to allow investigations of potentially non-linear 
relationships between an exposure and an outcome via a Mendelian randomization
framework, without requiring full access to individual level genetic data.

It is based on the existing package for individual data by James Staley: nlmr 
(available from  https://github.com/jrs95/nlmr ). 

The core concept is to split the process into two distinct halfs: one requiring 
individual level data, which is converted into a semi-summarized form 
(create_nlmr_summary) by dividing the population into strata based on the 
IV-free exposure. Associations with the exposure and the outcome are estimated 
in each stratum.
In the second half, this semi-summarized form can then be shared, without 
compromising patient privacy, and investigated seperately using two IV methods:
a fractional polynomial method (frac_poly_summ_mr) and a piecewise linear method
(piecewise_summ_mr). Both methods calculate a localised causal effect (LACE). 
The piecewise method fits a continuous piecewise linear function to these 
estimates, while the fractional polynomial method fits the best 1 or 2 term 
fractional polynomial.


## Functions
*create_nlmr_summary* - prepares individual level data into semi-summarised form,
ready to fit nlmr models.
*fracpoly_summ_mr* - this method performs IV analysis using fractional 
polynomials
*piecewise_summ_mr* - this method performs IV analysis using piecewise linear 
function

## Installation

You can install the released version of SUMnlmr from 
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("amymariemason/SUMnlmr")
```
## Example 1: Summarizing data

This is a basic example which shows you how to create the semi-summarized data 
form. First we create some practise data:

```{r example}
library(SUMnlmr)
## create some data to practise on
test_data<-create_ind_data(N=10000, beta2=2, beta1=1)
# this creates quadratic.Y  = x + 2x^2 + errorY 
head(test_data)
```

Then we use create_nlmr_summary to summarise it.

```{r example2}

## create the summarized form
## 
summ_data<-create_nlmr_summary(y = test_data$quadratic.Y,
                                x = test_data$X,
                                g = test_data$g,
                                covar = NULL,
                                family = "gaussian",
                                strata_method = "residual", 
                                controlsonly = FALSE,
                                q = 10)

head(summ_data$summary)
```
If we have co-variants we want to adjust for in our analysis, we need to include
them at this stage. 

```{r example3}

## create the summarized form
summ_covar<-create_nlmr_summary(y = test_data$quadratic.Y,
                                x = test_data$X,
                                g = test_data$g,
                                covar = matrix(data=c(test_data$linear.Y,
                                                      test_data$sqrt.Y),ncol=2),
                                family = "gaussian",
                                strata_method = "residual", 
                                q = 10)

head(summ_covar$summary)
```
Note: Because the covariants are included as a matrix, lm cannot detect factor 
variables and create automatic dummy variables for them. The easiest way to 
include factor variables is to make these dummy variables by hand instead using 
the [model.matrix](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/model.matrix.html)
command. e.g.

```{r example3b}

## create a factor
test_data$centre<- as.factor(rbinom(nrow(test_data),4, 0.5))

#turn factor into binary contrasts against first factor
dummies<- model.matrix(~centre,data=test_data)[,2:5]

summ_covar2<-create_nlmr_summary(y = test_data$quadratic.Y,
                                 x = test_data$X,
                                 g = test_data$g,
                                 covar = dummies,
                                 family = "gaussian",
                                 q = 10)

head(summ_covar2$summary)
```

These have used a single genetic variant count, but the method works identically 
with an genetic score function for g instead. Logistic or cox models in the G-Y relationship can be used by changing the family option - see details in the create_nlmr_summary function description.  

It is also possible to implement the doubly-ranked method described in Haodong's paper https://www.biorxiv.org/content/10.1101/2022.06.28.497930v1 


```{r example3c}

## create the summarized form with the doubly ranked method
summ_ranked<-create_nlmr_summary(y = test_data$quadratic.Y,
                                x = test_data$X,
                                g = test_data$g,
                                covar = matrix(data=c(test_data$linear.Y,
                                                      test_data$sqrt.Y),ncol=2),
                                family = "gaussian",
                                strata_method = "ranked", 
                                q = 10)

head(summ_ranked$summary)
```

Once your data is in this format, the output data frame is all you need to share
to fit the fractional polynomial or piecewise linear models onto the data.

## Example 2: Fitting a fractional polynomial model

Your data needs to be in the semi-summarised form as shown above. We can then
fit a fractional polynomial model:

```{r example4}


model<- with(summ_data$summary, frac_poly_summ_mr(bx=bx,
                  by=by, 
                  bxse=bxse, 
                  byse=byse, 
                  xmean=xmean,
                  family="gaussian",
                  fig=TRUE)
)


summary(model)


```
This also produces a graph of the fit with 95% confidence intervals.
This is a ggplot object and can be adjusted with ggplot commands
```{r example5}
library(ggplot2)
f <- function(x) (x + 2*x^2 - mean(summ_data$summary$xmean) -
                    2*mean(summ_data$summary$xmean)^2 )

plot1 <- model$figure+ 
  stat_function(fun = f, colour = "green") +
  ggtitle("fractional polynomial fit from semi-summarized data")

plot1
```
There is also p-values provided in p_test and p_het. This is identical to the 
testing provided by the nlmr package: 
* fp_d1_d2 :  test between the fractional polynomial degrees
* fp : fractional polynomial non-linearity test 
* quad: quadratic test
* Q : Cochran Q test
and 
* Q: Cochran Q heterogeneity test
* trend: trend test

``` {r example6}
model$p_tests

model$p_heterogeneity
```

## Example 3: Piecewise linear model

We can instead fit a piecewise linear model to the same summarised data

```{r plexample}

model2 <-with(summ_data$summary, piecewise_summ_mr(by, bx, byse, bxse, xmean, xmin,xmax, 
                  ci="bootstrap_se",
                  nboot=1000, 
                  fig=TRUE,
                  family="gaussian",
                  ci_fig="ribbon")
)

summary(model2)


```
Again the figure is a ggplot object and can be adjusted similarly. 

```{r pl2}
plot2 <- model2$figure+ 
  stat_function(fun = f, colour = "green") +
  ggtitle("piecewise linear fit from semi-summarized data")

plot2
```

## Example 4: Binary outcome

The functions above can also fit binary outcome data, via a generalised linear 
model. 

``` {r bin}

test_data$y.bin<-stats::rbinom(size=1, p=0.5, n=10000)

# create summ data
summ_bin<-create_nlmr_summary(y = test_data$y.bin,
                                x = test_data$X,
                                g = test_data$g,
                                covar = NULL,
                                family = "binomial",
                                q = 10)

# fit fractional poly model


model3<- with(summ_bin$summary,frac_poly_summ_mr(bx=bx,
                  by=by, 
                  bxse=bxse, 
                  byse=byse, 
                  xmean=xmean,
                  family="binomial",
                  fig=TRUE)
)

summary(model3)


```
Not unsurprisingly, we find no evidence of an effect, causal or otherwise, 
as the binary outcome was randomly distributed. 

If we look instead at the semi-summarised UK Biobank datasets on LDL-cholesterol 
and CAD, one with and one without covariates. Here we can see a potentially 
non-linear trend in the univariate data, which becomes a clear linear trend once
covariates are included.

```{r bi2}

# fit piecewise linear model
model4 <-with(LDL_CAD, piecewise_summ_mr(by, bx, byse, bxse, xmean, xmin,xmax, 
                  ci="bootstrap_se",
                  nboot=1000, 
                  fig=TRUE,
                  family="gaussian",
                  ci_fig="ribbon")
)


summary(model4)


# fit piecewise linear model
model5 <-with(LDL_CAD_covar,piecewise_summ_mr(by, bx, byse, bxse, xmean, xmin,xmax, 
                  ci="bootstrap_se",
                  nboot=1000, 
                  fig=TRUE,
                  family="gaussian",
                  ci_fig="ribbon")
)

summary(model5)

```
